embedplyr: Tools for Working With Text Embeddings

### 2025-02-11 paper/paper.Rmd
shorten paper

# Aims of the Package

@@

One particularly useful feature of `load_embeddings()` is the optional `words` parameter, which allows the user to specify a subset of words to load from the model. This allows users to work with large models, which are often too large to load into an interactive environment in their entirety. For example, a user may specify the set of unique tokens in their corpus of interest, and load only these from the model.

@@

DDR is ideal for studies of abstract constructs like emotions, that refer to the general gist of a text rather than particular words. The rich representation of word embeddings allows DDR to capture even the subtlest associations between words and constructs, and to precisely reflect the *extent* to which each word is associated with each construct. It can do this even for texts that do not contain any dictionary words. Because embeddings are continuous and calibrated by design to the probabilities of word use in language, DDR also avoids difficult statistical problems that arise due to the strange distributions of word counts [see @teitelbaum2024, chap. 16].

@@

## Other Functions

### Read Embeddings From File

Embeddings can come from many sources. Accordingly, **embedplyr** provides versatile tools to read embeddings from a variety of different formats. For example, @meng2019 provide spherical text embeddings pretrained on Wikipedia, formatted as text files. After downloading one of these files, it can be loaded using `read_embeddings()`. `read_embeddings()` can read GloVe text format, fastText text format, word2vec binary format, and most tabular formats (csv, tsv, etc.).

```{r, eval=FALSE}
spherical_mod <- read_embeddings("~/Downloads/jose_50d.txt")
```

Various R data types can also be converted to embeddings objects using `as.embeddings()`.

### Arbitrary Embedding Functions

In addition to embedding documents by averaging static word embeddings, `embed_docs()` can be used to wrap arbitrary embedding functions. For example, the **text** package [@kjell2023] can be used to generate embeddings using any model available from Hugging Face Transformers [@wolf2020]. This ability to wrap arbitrary functions is useful for analyses which require multiple corpora to be embedded in the same manner.

```{r, eval=FALSE}
# add embeddings to data frame
valence_sbert_df <- valence_df |> 
	embed_docs(
		"text", text::textEmbed, id_col = "id", .keep_all = TRUE,
		model = "sentence-transformers/all-MiniLM-L12-v2"
		)
```

### Quanteda Compatibility

The examples above used a **tidyverse** approach with tibbles as the primary data structure, but **embedplyr** functions can be used similarly within a **quanteda** workflow. In the example below, I reproduce the text embedding shown above using `textstat_embedding()` (following **quanteda** naming standards) as opposed to `embed_docs()`.

```{r, warning=FALSE, message=FALSE}
library(quanteda)

# corpus
psych_embeddings_corp <- corpus(psych_embeddings_df, docid_field = "id")

# dfm
psych_embeddings_dfm <- psych_embeddings_corp |> 
	tokens() |> 
	dfm()

# compute embeddings
psych_embeddings_df <- psych_embeddings_dfm |> 
	textstat_embedding(glove_twitter_25d)
```

### Similarity Metrics

Functions for similarity and distance metrics are as simple as possible; each one takes in vectors and outputs a scalar. In addition to cosine similarity, Euclidean distance, and dot product, **embedplyr** functions also support projection to anchored vectors [@grand2022], as shown in the examples above.

```{r, eval=FALSE}
dot_product <- dot_prod(vec1, vec2)
cosine_similarity <- cos_sim(vec1, vec2)
euclidean_distance <- euc_dist(vec1, vec2)
semantic_projection <- anchored_sim(vec1, pos = vec2, neg = vec3)
```

### Reduce Dimensionality

It is sometimes useful to reduce the dimensionality of embeddings. This is done with `reduce_dimensionality()`, which by default performs PCA without column normalization. Reducing the dimensionality to two or three dimensions can be useful for visualization.

```{r}
valence_df_2d <- psych_embeddings_df |> 
	reduce_dimensionality(dim_1:dim_25, 2)
valence_df_2d
```

`reduce_dimensionality()` can also be used to apply the same rotation to other embeddings not used to find the principle components.

```{r}
new_embeddings <- predict(glove_twitter_25d, c("new", "strange"))

# get rotation with `output_rotation = TRUE`
valence_rotation_2d <- psych_embeddings_df |> 
	reduce_dimensionality(dim_1:dim_25, 2, output_rotation = TRUE)

# apply the same rotation to new embeddings
new_with_valence_rotation <- new_embeddings |> 
	reduce_dimensionality(custom_rotation = valence_rotation_2d)
new_with_valence_rotation
```

### Align and Compare Embeddings Models

Aligning separately trained embeddings can be useful for tracking semantic change over time or semantic differences between training datasets. It can also be useful for comparing texts across languages. `align_embeddings(x, y)` rotates the embeddings in `x` (and changes their dimensionality if necessary) so that they can be compared with those in `y` [@schonemann1966]. Optionally, `matching` can be used to specify one-to-one matching between  embeddings in the two models (e.g. a bilingual dictionary).

Once aligned, groups of embeddings can be compared using `total_dist()` or `average_sim()`. These can be used to quantify the overall distance or similarity between two parallel embedding spaces.

### Normalize (Scale Embeddings to the Unit Hypersphere)

`normalize()` and `normalize_rows()` scale embeddings such that their magnitude is 1, while their angle from the origin is unchanged.

```{r}
normalize(1:10)

normalize(moral_embeddings)

psych_embeddings_normalized <- psych_embeddings_df |>
	normalize_rows(dim_1:dim_25)
```

### Magnitude

The magnitude, norm, or length of a vector is its Euclidean distance from the origin.

```{r}
magnitude(1:10)

magnitude(moral_embeddings)
```

### Similarity Matrix

Given an embeddings object or data frame of embeddings, `sim_matrix()` calculates a matrix of similarity scores between the rows of the input. This can be useful for analyzing distributional properties of groups of documents [see e.g. @markus2024] or of groups of tokens [see e.g. @choi2024].

```{r, warning=FALSE}
text <- c("lions and tigers and bears", "curious green ideas")

text |> 
	embed_tokens(glove_twitter_25d, output_embeddings = TRUE) |> 
	sim_matrix(method = "euclidean")
```

>>>>

# Statement of Need

@@

One particularly useful feature of `load_embeddings()` is the optional `words` parameter, which allows the user to specify a subset of words to load from the model. This allows users to work with large models, which are often too large to load into an interactive environment in their entirety.

@@

@@
